Computing solution space properties by generic programming tensor networks \thanks{\funding{...}}

### 2022-03-31 paper/paper.tex
Updates from Overleaf

\begin{abstract}
%We introduce a method that uses generic programming tensor networks to compute various properties of a class of hard graph problems that can be decribed by local constraints.
We introduce a tensor network algorithm %a method that combines generic programming and tensor network techniques 
to compute various solution space properties of a class of combinatorial optimization problems including the independent set problem, the maximum cut problem, the vertex coloring problem, the maximal clique problem, the dominating set problem, and the satisfiability problem, among others.
%Let the independent set problem be an example, we show how to compute, for example, the maximum several independent set sizes, the counting of independent sets of a give size, the enumeration of independent sets of a given size.
We look at the independent set problem as an example, and show how to compute the size of the maximum independent set, count the number of independent sets of a given size, and enumerate/sample the independent sets of a given size.
%By making use of generic programming, our algorithms are very simple to implement and one can in addition directly utilize recent advances in tensor network contraction techniques such as near-optimal contraction order finding and slicing to achieve high performance.
By using generic programming techniques, the same framework can be used to compute all of these properties. Our algorithm utilizes recent advances in tensor network contraction techniques to achieve high performance, including methods to quickly find a near-optimal contraction order and slicing.
%The algorithmic complexity of this approach is two to the treewidth of the line graph of the tensor network graph.
To demonstrate how our versatile tool helps to understand these hard problems, we apply it to a few examples, including computing the entropy constant for several hardcore gases on disordered lattices, studying the overlap gap property on different classes of graphs, and analyzing the performance of quantum and classical optimization algorithms for the independent set problem.% including the calculations of the entropy constant for some hardcore lattice gases on 2D square lattices and the study of the overlap gap property on three regular graphs.
%We show how our method helps to understand these hard problems on systems with hundreds of vertices.
\end{abstract}

@@

\section{Tensor networks}\label{sec:tn}
A tensor network~\cite{Cirac2020, Orus2014} is composed of a collection of tensors, a collection of labels associated with tensor dimensions and an indicator of which tensor is the output.
The operation to evaluate a tensor network is called \textit{contraction}, which is defined as a summation of products of tensor elements over the labels not appearing in the output tensor.
For example, the matrix multiplication is a special tensor network that can be represented as $C_{ik} = A_{ij}B_{jk}$, where $A, B$ and $C$ are tensors, $ik,ij$ and $ik$ are labels associated with them, and ``$=$'' is the indicator of $C$ being the output, while its contraction is defined as $C_{ik} = \sum_j A_{ij}B_{jk}$.
%can be viewed as a generalization of binary matrix multiplication to nary tensor contraction.
%Generalized Einstein's notation is often used to represent a tensor network, where we use a label in the subscript to represent a degree of freedom.
%One can enumerate these degrees of freedom and accumulate the product of tensor elements to the output tensor.
% Einstein's notation is often used to represent a tensor network, e.g.\ it represents the matrix multiplication between two matrices $A$ and $B$ as $C_{ik} = A_{ij}B_{jk}$,
% where we use a label in the subscript to represent a degree of freedom.
% One can enumerate these degrees of freedom and accumulate the product of tensor elements to the output tensor.
% In the standard Einstein's notation for tensor networks in physics, each index appears precisely twice, either both in input tensors (which will be summed over) or one in a input tensor and another in the output tensor.
% Hence a tensor network can be represented as a open simple graph,
% where a tensor is mapped to a vertex, a label shared by two input tensors is mapped to a normal edge and a label appears in the output tensor is mapped to an open edge.
% In this work, we adopt a more general form of tensor networks that do not restrict how many times a label can appear in tensors. 
The graphical representation of a tensor network is a open (meaning a hyperedge can connect to external vertices) hypergraph, where an input tensor is mapped to a vertex and a label is mapped to a hyperedge that can connect an arbitrary number of vertices, while the labels appearing in the output tensor are open.
Our notation slightly generalizes the standard tensor network used in physics by not restricting how many times a label can appear in tensors. 
While this generalized form is equivalent in representation power, it can have smaller contraction complexity as will be illustrated in \App{sec:tensorbad}.
% because one can easily translate a generalized tensor network to the standard notation by adding $\delta$ tensors, a high dimensional equivalence of the identity matrix.
%However, introducing $\delta$ tensors may significantly increase the complexity of the tensor contraction. We illustrate this subtle point in \App{sec:tensorbad}.

@@

\section{Generic programming tensor contractions}\label{sec:generic}
%\purple{On a second read of this section, I think that the clarity could be improved. My understanding is this is the flow: We would like to use generic programming, where we write code that works regardless of input type, and is optimized on all input types without sacrificing efficiency. In general, we can write code such that a function runs correctly regardless of the input data type. This is easy to do for dynamically typed languages like Python, which only verify and enforce the rules for the input data type at run time (called type-checking). However, when the input data type is only checked at run time (as opposed to when the code is compiled), type-specific optimization cannot be used. High efficiency is only achieved in statically typed languages, where modern compiling technology allows the function to be individually optimized for each input data type (then give C++/Julia examples).}\blue{how about now?}
In previous works relating tensor networks and combinatoric problems~\cite{Kourtis2019, Biamonte2017},
the elements in the tensor networks are limited to standard number types such as floating point numbers and integers.
%Owning to the development of modern compiling technology, we no longer need to restrict our imagination to standard number types.
However, the elements in a tensor do not have to be a regular number type,
 or even a field (since division is not used in tensor contraction).
One can use the same tensor contraction program for different purposes with different tensor element types.
This idea of using the same program for different purposes is also called the generic programming in computer science:

\begin{definition}[Generic programming ~\cite{Stepanov2014}]
   Generic programming is an approach to programming that focuses on designing algorithms and data structures so that they work in the most general setting without loss of efficiency.
\end{definition}
This definition of generic programming covers two major aspects: ``work in the most general setting'' and ``with out loss of efficiency''.
By in the most general setting, we mean that a single program should work correctly for different input data types. For example, 
%\purple{I think the following paragraph could be clarified. When you read the definition of generic programming, it's not clear what ``in the most general setting'' means. My understanding is it just means that the program will work regardless of the input data type - it would be good to say this explicitly. }
%\blue{This definition is from the book I cited, I also do not like ``in the most general setting'', maybe we just use our version?}
%\purple{I also think the way it is written takes a while to understand what dynamically versus statically typed languages are. The definition, “When these languages “see” a new input type, the compiler can recompile the generic program for the new type.” comes rather late in the paragraph.}
%\xpurple{Also, while I know what you mean, concepts like “inlining immutable elements” and “cache miss rate” would be
%lost on someone only with a physics background.}
%This definition of generic programming contains two major aspects: a single program works in the most general setting and efficiency.
%To understand the first aspect on generality,
suppose we want to write a function that raises an element to a power, $f(x, n) := x^n$.
One can easily write a function for standard number types that computes the power of $x$ in $O \left( \log(n) \right)$ steps using the multiply and square trick.
Generic programming does not require $x$ to be a standard number type,
instead it treats $x$ as an element with an associative multiplication operation $\odot$ and a multiplicative identity $\mymathbb{1}$.
Then, when the program takes a matrix as an input instead of a standard number type, it computes the matrix power correctly without rewriting the program.
The second aspect is about efficiency. For dynamically typed languages such as Python, the type information is not available for type specific optimization at the compiling stage.
Therefore, one can easily write code that works for general input types, but the efficiency is not guaranteed; for example, the speed of computing the matrix multiplication between two NumPy arrays with Python objects as elements is much slower than statically typed (i.e. the type information is used in the compiling stage) languages such as C++ and Julia~\cite{Bezanson2012}.
C++ uses templates for generic programming while Julia takes advantage of just-in-time compilation and multiple dispatch.
When these languages ``see'' a new input type, the compiler can recompile the generic program for the new type.
A myriad of optimizations can be done during the compilation. For example, the compiler can optimize the memory layout of immutable elements with fixed sizes in an array to speed up the array indexing.
In Julia, if a type is immutable and contains no references to other values, an array of that type can even be compiled to graphics processing units (GPU) for faster computation~\cite{Besard2018}.

This motivates us to identify the most general tensor element type allowed in a tensor network contraction program.
We find that as long as the tensor elements are members of a commutative semiring, the tensor network contraction will be well defined and the result will be independent of the contraction order.
Here, a commutative semiring is a field without additive inverse and multiplicative inverse.
Giving up these nice properties of  fields has significant implications for tensor computation: tensor network compression algorithms might be not applicable because matrix factorization is NP-hard for commutative semirings~\cite{Shitov2014}, and matrix multiplication faster than $O(n^3)$ does not exist for an algebra without an additive inverse~\cite{Kerr1970}.
But here, we only use the commutative properties of an algebra for optimizing the tensor network contraction order.
To define a commutative semiring with the addition operation $\oplus$ and the multiplication operation $\odot$ on a set $S$, the following relations must hold for any arbitrary three elements $a, b, c \in S$.
\begin{align*}
(a \oplus b) \oplus c = a \oplus (b \oplus c) & \hspace{5em}\text{$\triangleright$ commutative monoid $\oplus$ with identity $\mymathbb{0}$}\\
a \oplus \mymathbb{0} = \mymathbb{0} \oplus a = a &\\
a \oplus b = b \oplus a &\\
&\\
(a \odot b) \odot c = a \odot (b \odot c)  &   \hspace{5em}\text{$\triangleright$ commutative monoid $\odot$ with identity $\mymathbb{1}$}\\
a \odot  \mymathbb{1} =  \mymathbb{1} \odot a = a &\\
a \odot b = b \odot a &\\
&\\
a \odot (b\oplus c) = a\odot b \oplus a\odot c  &  \hspace{5em}\text{$\triangleright$ left and right distributive}\\
(a\oplus b) \odot c = a\odot c \oplus b\odot c &\\
&\\
a \odot \mymathbb{0} = \mymathbb{0} \odot a = \mymathbb{0}
\end{align*}
%We require the algebra being multiplicative commutative because the contraction order optimizer will change the multiplication order of elements, while we require that the contraction order does not change the contraction results.
In the following sections, we show how to compute a number of properties of independent sets by designing tensor element types as specific commutative semirings without changing the tensor network contraction program~\cite{Stepanov2014}.
The Venn diagram in \Fig{fig:venn-diagram} shows the algebras we will introduce in the main text and their relation, and Table~\ref{tbl:generictypes} summarizes which properties can be computed by which tensor element types.

@@

    \caption{The the tensor network element types used in this work, while the purpose of these element types can be found in \Tbl{tbl:generictypes}.
    The overlap between two algebra indicates that a new algebra can be created by combining those two types of algebra. ``Largest order'' and ``Largest 2 orders'' mean truncating the polynomial by only keeping its largest or largest two orders.}
     \label{fig:venn-diagram}
\end{figure}

@@

\section{Tensor network representation of independent sets} \label{sec:tnmap}
This section is about reducing the independent set problem to a tensor network contraction problem.
An alternative approach dedicated to physicists can be found in \App{sec:energymodel}, where we introduce the reduction from the energy model of hard core lattice gases~\cite{Dyre2016, Fernandes2007}.
Let $G=(V,E)$ be an undirected graph with each vertex $v\in V$ being associated with a weight $w_v$.
An independent set $I \subseteq V$ is a set of vertices that for any vertex pair $u,v \in I$, we have $(u, v) \not\in E$.
In its tensor network representation, a vertex $v\in V$ is mapped to a label $s_v \in \{0, 1\}$ of dimension $2$, where we use $0$ ($1$) to denote this vertex is absent (present) in the set.
For each label $s_v$, we defined a parametrized rank-one vertex tensor $W(x_v^{w_v})$ as
\begin{equation}
    W(x_v^{w_v}) = \left(\begin{matrix}
        1 \\
        x_v^{w_v}
    \end{matrix}\right),\label{eq:vertextensor}
\end{equation}
where $x_v$ is a variable associated with vertex $v$.
Tensor elements can be indexed by subscripts, e.g.\ $W(x_v^{w_v})_0=1$ is the first element associated with $s_v=0$ and $W(x_v^{w_v})_1=x_v^{w_v}$ is the second element associated with $s_v=1$.
Similarly, on each edge $(u, v)$, we define a matrix $B$ indexed by $s_u$ and $s_v$ as
\begin{equation}
    \qquad \quad 
       B = \left(\begin{matrix}
        1  & 1\\
        1 & 0
    \end{matrix}\right). \label{eq:edgetensor}
\end{equation}
The corresponding tensor network contraction is defined as
\begin{equation}\label{eq:idp}
    P(G, \{x_1^{w_1}, x_{2}^{w_2}, \ldots,x_{|V|}^{w_{|V|}}\}) = \sum\limits_{s_1, s_2, \ldots, s_{|V|} = 0}^{1} \prod\limits_{v\in V} W(x_v^{w_v})_{s_v} \prod\limits_{(u,v) \in E} B_{s_u s_v},
\end{equation}
where the summation runs over all vertex configurations $\{s_1, s_{2}, \ldots,s_{|V|}\}$ and accumulates the product of tensor elements to the output $P$.
An edge tensor element $B_{s_us_v}$ encodes the independent set constraint, which zero out the product whenever $u$ and $v$ are both in the independent set.
While a vertex tensor element $W(x_v^{w_v})_{s_v}$ contributes a multiplicative factor $x_v^{w_v}$ if the vertex $v$ is in the independent set.

@@

In general, it is not computationally efficient to evaluate \Eq{eq:idp} by directly summing up the $2^{|V|}$ products. %, which scales exponentially as the number of vertices.
%\purple{Perhaps explicitly say what this scales as?}\blue{I am not sure explicitly saying ``scales exponentially as the number of vertices'' is helpful, because treewidth often scales linearly as number of vertices, so tensor network also scales exponentially as the number vertices in most cases.}
The standard approach to evaluate a randomly structured tensor network is: first find a good pair-wise tensor contraction order as a binary tree, and then contract two tensors a time by this order.
An optimal tensor network contraction order corresponds to an optimal tree decomposition of its line graph (here it is equivalent to the graph this tensor mapped from) and the largest intermediate tensor has a rank equal to the tree width of this optimal tree decomposition~\cite{Markov2008}.
In practice, it is difficult to find an optimal contraction order for large tensor networks because finding the treewidth itself is NP-hard.
However, it is easy to find a close to optimal contraction order in typically a few minutes with a heuristic algorithm~\cite{Kourtis2019,Kalachev2021}.
%The pairwise tensor contraction also makes it possible to utilize basic linear algebra subprograms (BLAS) functions to speed up the computation for certain tensor element types.
For large scale applications, it is also possible to slice over certain degrees of freedom to reduce the space complexity, i.e.
loop and accumulate over certain degrees of freedom so that one can have a smaller tensor network inside the loop due to the removal of these degrees of freedom.

@@

\section{Discussion and conclusion}
In this paper, we introduce using generic tensor networks to compute a number of different solution space properties of a certain class of NP-hard combinatorial optimization problems. % the maximum independent set problem. %, 
These solution space properties include the best several solutions, the number of solution configurations of a given size, and the enumeration and sampling of solution configurations of a given size.
For each solution space property, we compute it with the same tensor network by replacing the tensor elements with those having an algebra designed for it. % This technique can further be generalized to computing solution space properties beyond the examples given in this paper. % such as the largest several independent sets sizes, the number of independent sets of a given size, and the enumeration of independent sets of a given size.
%With different algebras, the contraction of the same tensor network gives us the different properties.
The different data types introduced in the main text to compute these properties are  summarized in the diagram in \Fig{fig:venn-diagram}. %We go beyond using real numbers and use polyn
Other than the case with real numbers, we have polynomials and truncated polynomials.
We combine them with the bit string algebra, the set algebra, and sum-product expression tree algebra for finding, enumerating, and sampling of independent sets at a given size. Three examples are shown to demonstrate how our method can be used to understand the solution space better.
In the first example, we show how our method can help computing the entropy constant for some hardcore lattice gases on 2D square lattices. We compute the 2D generalization of Fibonacci integer sequence up to size $39$, which is larger than the previous record $37$.
In the second example, we analyze the pairwise Hamming distances from configurations directly sampled from large independent sets of King's graphs at $0.8$ filling and three regular graphs. We show the strongest numeric evidence so far of the absence of overlap gap property in King's graph and the existence of Overlap Gap properties in three regular graphs.
In the last example, we show how our algorithm helps people to understand the performance of a quantum variational algorithms on Rydberg atom arrays to find maximum independent sets. This is a research frontier that understanding which can help people design better quantum algorithms to beat its classical counterpart.
These algebras are of generic use: they can be utilized to compute properties of maximal independent sets and a variety of other combinatorial problems such as the matching problem, the k-coloring problem, the max-cut problem, and the set packing problem, as detailed in \App{sec:otherproblems}.

It is worth mentioning that a tensor network is very related to a well known computer programming method, the dynamic programming~\cite{Courcelle1990, Fomin2013}, like the Viterbi algorithm for finding most probable configuration in the hidden Markov model has a nice matrix product state interpretation~\cite{Han2018}, and the tropical tensor network is very likely equivalent to dynamic programming in solving hard combinatorial optimization problems in certain way~\cite{Liu2021} and they have the same time complexity related to the treewidth of a graph.
However, while being more restrictive regarding the problem set it can solve, the tensor network has a richer algebraic structure, which crucially enables the computation of solution space properties beyond finding a single solution, which is not possible with dynamic programming.
%Moreover, since the independence polynomial is closely related to the matching polynomial~\cite{Levit2005}, the clique polynomial~\cite{Hoede1994}, and the vertex cover polynomial~\cite{Akbari2013},
%our algorithm to compute the independence polynomial can also be used to compute these graph polynomials.
%By adapting to the style of generic programming, we can implement all the algorithms without much effort.

The source code in Julia language for this paper can be found in our Github repository~\cite{GraphTensorNetworks}. 
There is also a short introduction to this repository as well a gist to show how it works in the Appendix~\ref{sec:technical}.
We expect our tool can be used to understand and study many interesting applications of independent sets and beyond. We also hope this approach of generic tensor networks can inspire future works on tensor network computations.

>>>>

\begin{abstract}
%We introduce a method that uses generic programming tensor networks to compute various properties of a class of hard graph problems that can be decribed by local constraints.
%We introduce a tensor network algorithm %a method that combines generic programming and tensor network techniques 
We discuss computing solution space properties of a class of combinatorial optimization problems including the independent set problem, the maximum cut problem, the vertex coloring problem, the maximal clique problem, the dominating set problem, and the satisfiability problem, among others, where
%Let the independent set problem be an example, we show how to compute, for example, the maximum several independent set sizes, the counting of independent sets of a give size, the enumeration of independent sets of a given size.
the solution space properties include finding one of the best solution, counting the number of solutions of a given size, and enumeration and sampling of solutions of a given size.
We look at the independent set problem as an example, and show how all these solution space properties can be solved in a unified framework of generic programming tensor networks.
%By making use of generic programming, our algorithms are very simple to implement and one can in addition directly utilize recent advances in tensor network contraction techniques such as near-optimal contraction order finding and slicing to achieve high performance.
%By using generic programming techniques, the same framework can be used to compute all of these properties. 
%Our algorithm utilizes recent advances in tensor network contraction techniques to achieve high performance, including methods to quickly find a near-optimal contraction order and slicing.
%The algorithmic complexity of this approach is two to the treewidth of the line graph of the tensor network graph.
To demonstrate how our versatile tool helps to understand these hard problems, we apply it to a few examples, including computing the entropy constant for hardcore lattice gases, studying the overlap gap properties, and analyzing the performance of quantum and classical algorithms for finding maximum independent sets.% including the calculations of the entropy constant for some hardcore lattice gases on 2D square lattices and the study of the overlap gap property on three regular graphs.
%We show how our method helps to understand these hard problems on systems with hundreds of vertices.
\end{abstract}

@@

\section{Tensor networks}\label{sec:tn}
A tensor network~\cite{Cirac2020, Orus2014} is composed of a collection of tensors, a collection of labels associated with tensor dimensions and an indicator of which tensor is the output.
The operation to evaluate a tensor network is called \textit{contraction}, which is defined as a summation of tensor element products over the labels not appearing in the output tensor.
For example, the matrix multiplication is a special tensor network that can be represented as $C_{ik} = A_{ij}B_{jk}$, where $A, B$ and $C$ are tensors, $ik,ij$ and $ik$ are labels associated with them, and ``$=$'' is the indicator of $C$ being the output, while its contraction is defined as $C_{ik} = \sum_j A_{ij}B_{jk}$.
%can be viewed as a generalization of binary matrix multiplication to nary tensor contraction.
%Generalized Einstein's notation is often used to represent a tensor network, where we use a label in the subscript to represent a degree of freedom.
%One can enumerate these degrees of freedom and accumulate the product of tensor elements to the output tensor.
% Einstein's notation is often used to represent a tensor network, e.g.\ it represents the matrix multiplication between two matrices $A$ and $B$ as $C_{ik} = A_{ij}B_{jk}$,
% where we use a label in the subscript to represent a degree of freedom.
% One can enumerate these degrees of freedom and accumulate the product of tensor elements to the output tensor.
% In the standard Einstein's notation for tensor networks in physics, each index appears precisely twice, either both in input tensors (which will be summed over) or one in a input tensor and another in the output tensor.
% Hence a tensor network can be represented as a open simple graph,
% where a tensor is mapped to a vertex, a label shared by two input tensors is mapped to a normal edge and a label appears in the output tensor is mapped to an open edge.
% In this work, we adopt a more general form of tensor networks that do not restrict how many times a label can appear in tensors. 
The graphical representation of a tensor network is an open (meaning an edge can connect to external vertices) hypergraph, where an input tensor is mapped to a vertex and a label is mapped to a hyperedge that can connect an arbitrary number of vertices, while the labels appearing in the output tensor are open.
Our notation slightly generalizes the standard tensor network used in physics by not restricting how many times a label can appear in tensors. 
While this generalized form is equivalent in representation power, it can have smaller contraction complexity as will be illustrated in \App{sec:tensorbad}.
% because one can easily translate a generalized tensor network to the standard notation by adding $\delta$ tensors, a high dimensional equivalence of the identity matrix.
%However, introducing $\delta$ tensors may significantly increase the complexity of the tensor contraction. We illustrate this subtle point in \App{sec:tensorbad}.

@@

\section{Generic programming tensor contractions}\label{sec:generic}
%\purple{On a second read of this section, I think that the clarity could be improved. My understanding is this is the flow: We would like to use generic programming, where we write code that works regardless of input type, and is optimized on all input types without sacrificing efficiency. In general, we can write code such that a function runs correctly regardless of the input data type. This is easy to do for dynamically typed languages like Python, which only verify and enforce the rules for the input data type at run time (called type-checking). However, when the input data type is only checked at run time (as opposed to when the code is compiled), type-specific optimization cannot be used. High efficiency is only achieved in statically typed languages, where modern compiling technology allows the function to be individually optimized for each input data type (then give C++/Julia examples).}\blue{how about now?}
In previous works relating tensor networks and combinatoric problems~\cite{Kourtis2019, Biamonte2017}, the elements types in the tensor networks are limited to standard number types such as floating point numbers and integers.
%Owning to the development of modern compiling technology, we no longer need to restrict our imagination to standard number types.
%However, the elements in a tensor do not have to be a regular number type, or even a field (since division is not used in tensor contraction).
%One can use the same tensor contraction program for different purposes with different tensor element types.
We propose to not restricting our imagination to standard number types, but to use a more general element type with certain algebraic property.
With different element types, we can solve different problems within the same framework.
This idea of using the same program for different purposes is also called the generic programming in computer science:

\begin{definition}[Generic programming ~\cite{Stepanov2014}]
   Generic programming is an approach to programming that focuses on designing algorithms and data structures so that they work in the most general setting without loss of efficiency.
\end{definition}
This definition of generic programming covers two major aspects: ``work in the most general setting'' and ``with out loss of efficiency''.
By in the most general setting, we mean that a single program should work correctly for different input data types. 
%\purple{I think the following paragraph could be clarified. When you read the definition of generic programming, it's not clear what ``in the most general setting'' means. My understanding is it just means that the program will work regardless of the input data type - it would be good to say this explicitly. }
%\blue{This definition is from the book I cited, I also do not like ``in the most general setting'', maybe we just use our version?}
%\purple{I also think the way it is written takes a while to understand what dynamically versus statically typed languages are. The definition, “When these languages “see” a new input type, the compiler can recompile the generic program for the new type.” comes rather late in the paragraph.}
%\xpurple{Also, while I know what you mean, concepts like “inlining immutable elements” and “cache miss rate” would be
%lost on someone only with a physics background.}
%This definition of generic programming contains two major aspects: a single program works in the most general setting and efficiency.
%To understand the first aspect on generality,
For example, suppose we want to write a function that raises an element to a power, $f(x, n) := x^n$.
One can easily write a function for standard number types that computes the power of $x$ in $O \left( \log(n) \right)$ steps using the multiply and square trick.
Generic programming does not require $x$ to be a standard number type,
instead it treats $x$ as an element with an associative multiplication operation $\odot$ and a multiplicative identity $\mymathbb{1}$.
Then, when the program takes a matrix as an input instead of a standard number type, it computes the matrix power correctly without rewriting the program.
The second aspect is about efficiency. For dynamically typed languages such as Python, the type information is not available for type specific optimizations at the compiling stage.
Therefore, one can easily write code that works for general input types, but the efficiency is not guaranteed; for example, the speed of computing the matrix multiplication between two NumPy arrays with Python objects as elements is much slower than statically typed (i.e. the type information can be accessed at the compiling stage) languages such as C++ and Julia~\cite{Bezanson2012}.
C++ uses templates for generic programming while Julia takes advantage of just-in-time compilation and multiple dispatch.
When these languages ``see'' a new input type, the compiler recompiles the generic program for the new type to generate efficient binary.
A myriad of optimizations can be done during the compilation. For example, the compiler can optimize the memory layout of immutable elements with fixed sizes in an array to speed up array indexing.
In Julia, if a type is immutable and contains no references to other values, an array of that type can even be compiled to graphics processing units (GPU) for faster computation~\cite{Besard2018}.

This motivates us to identify the most general tensor element type allowed in a tensor network contraction program.
We find that as long as the tensor elements are members of a commutative semiring, the tensor network contraction will be well defined and the result will be independent of the contraction order.
Here, a commutative semiring is a field without additive inverse and multiplicative inverse.
Giving up these nice properties of  fields has significant implications for tensor computation: tensor network compression algorithms might be not applicable because matrix factorization is NP-hard for commutative semirings~\cite{Shitov2014}, and matrix multiplication faster than $O(n^3)$ does not exist for an algebra without an additive inverse~\cite{Kerr1970}.
But here, we only use the commutative properties of an algebra for optimizing the tensor network contraction order.
To define a commutative semiring with the addition operation $\oplus$ and the multiplication operation $\odot$ on a set $S$, the following relations must hold for any arbitrary three elements $a, b, c \in S$.
\begin{align*}
(a \oplus b) \oplus c = a \oplus (b \oplus c) & \hspace{5em}\text{$\triangleright$ commutative monoid $\oplus$ with identity $\mymathbb{0}$}\\
a \oplus \mymathbb{0} = \mymathbb{0} \oplus a = a &\\
a \oplus b = b \oplus a &\\
&\\
(a \odot b) \odot c = a \odot (b \odot c)  &   \hspace{5em}\text{$\triangleright$ commutative monoid $\odot$ with identity $\mymathbb{1}$}\\
a \odot  \mymathbb{1} =  \mymathbb{1} \odot a = a &\\
a \odot b = b \odot a &\\
&\\
a \odot (b\oplus c) = a\odot b \oplus a\odot c  &  \hspace{5em}\text{$\triangleright$ left and right distributive}\\
(a\oplus b) \odot c = a\odot c \oplus b\odot c &\\
&\\
a \odot \mymathbb{0} = \mymathbb{0} \odot a = \mymathbb{0}
\end{align*}
%We require the algebra being multiplicative commutative because the contraction order optimizer will change the multiplication order of elements, while we require that the contraction order does not change the contraction results.
In the following sections, we will show how to compute a number of properties of independent sets using the same tensor network contraction algorithm by engineering tensor element algebras.
The Venn diagram in \Fig{fig:venn-diagram} shows the algebras we will introduce in the main text and their relation, and Table~\ref{tbl:generictypes} summarizes which properties can be computed by which tensor element types.

@@

    \caption{The the tensor network element types used in this work and their relationships.
    The overlap between two ellipses indicates that a new algebra can be created by combining those two algebras. ``Largest order'' and ``Largest 2 orders'' mean truncating the polynomial by only keeping its largest or largest two orders.
    The purposes of these element types can be found in \Tbl{tbl:generictypes}.}
     \label{fig:venn-diagram}
\end{figure}

@@

\section{Tensor network representation of independent sets} \label{sec:tnmap}
This section is about reducing the independent set problem to a tensor network contraction problem.
An alternative approach dedicated to physicists can be found in \App{sec:energymodel}, where we introduce the reduction from the energy model of hard core lattice gases~\cite{Dyre2016, Fernandes2007}.
Let $G=(V,E)$ be an undirected graph with each vertex $v\in V$ being associated with a weight $w_v$.
An independent set $I \subseteq V$ is a set of vertices that for any vertex pair $u,v \in I$, we have $(u, v) \not\in E$.
To reduce the independent set problem on $G$ to a tensor network contraction, we first map a vertex $v\in V$ to a label $s_v \in \{0, 1\}$ of dimension $2$, where we use $0$ ($1$) to denote a vertex absent (present) in the set.
For each vertex $v$, we defined a parameterized rank-one tensor indexed by $s_v$ as
\begin{equation}
    W(x_v^{w_v}) = \left(\begin{matrix}
        1 \\
        x_v^{w_v}
    \end{matrix}\right),\label{eq:vertextensor}
\end{equation}
where $x_v$ is a variable associated with $v$.
Tensor elements can be indexed by subscripts, e.g.\ $W(x_v^{w_v})_0=1$ is the first element associated with $s_v=0$ and $W(x_v^{w_v})_1=x_v^{w_v}$ is the second element associated with $s_v=1$.
Similarly, for each edge $(u, v) \in E$, we define a matrix $B$ indexed by $s_u$ and $s_v$ as
\begin{equation}
    \qquad \quad 
       B = \left(\begin{matrix}
        1  & 1\\
        1 & 0
    \end{matrix}\right). \label{eq:edgetensor}
\end{equation}
The corresponding tensor network contraction is defined as
\begin{equation}\label{eq:idp}
    P(G, \{x_1^{w_1}, x_{2}^{w_2}, \ldots,x_{|V|}^{w_{|V|}}\}) = \sum\limits_{s_1, s_2, \ldots, s_{|V|} = 0}^{1} \prod\limits_{v\in V} W(x_v^{w_v})_{s_v} \prod\limits_{(u,v) \in E} B_{s_u s_v},
\end{equation}
where the summation runs over all $2^{|V|}$ vertex configurations $\{s_1, s_{2}, \ldots,s_{|V|}\}$ and accumulates the product of tensor elements to the output $P$.
It is easy to see an edge tensor element $B_{s_us_v}$ encodes the independent set constraint. It zeros out the product whenever both $u$ and $v$ are in the independent set.
Meanwhile, a vertex tensor element $W(x_v^{w_v})_{s_v}$ contributes a multiplicative factor $x_v^{w_v}$ whenever $v$ is in the set.

@@

For a general graph, it is computationally inefficient to evaluate \Eq{eq:idp} by directly summing up the $2^{|V|}$ products. %, which scales exponentially as the number of vertices.
%\purple{Perhaps explicitly say what this scales as?}\blue{I am not sure explicitly saying ``scales exponentially as the number of vertices'' is helpful, because treewidth often scales linearly as number of vertices, so tensor network also scales exponentially as the number vertices in most cases.}
The standard approach to evaluate a randomly structured tensor network is: first find a good pair-wise tensor contraction order as a binary tree, and then contract two tensors a time by this order.
A contraction order of a tensor network corresponds to a tree decomposition of the line graph of its hypergraph representation and the largest intermediate tensor has a rank equal to the width of this tree decomposition~\cite{Markov2008}.
An optimal (in terms of space complexity) contraction order corresponds to the tree decomposition with smallest width, i.e. the treewidth.
Here, this line graph is equivalent to the graph this tensor network mapped from. 
In practice, it is difficult to find an optimal contraction order for large tensor networks because finding the treewidth itself is NP-hard.
However, it is easy to find a close to optimal contraction order within typically a few minutes with a heuristic algorithm~\cite{Kourtis2019,Kalachev2021}.
%The pairwise tensor contraction also makes it possible to utilize basic linear algebra subprograms (BLAS) functions to speed up the computation for certain tensor element types.
For large scale applications, it is also possible to slice over certain degrees of freedom to reduce the space complexity, i.e.
loop and accumulate over certain degrees of freedom so that one can have a smaller tensor network inside the loop due to the removal of these degrees of freedom.

@@

\section{Discussion and conclusion}
In this paper, we introduce using generic tensor networks to compute a number of different solution space properties of a certain class of NP-hard combinatorial optimization problems. % the maximum independent set problem. %, 
These solution space properties include the best several solutions, the number of solutions of a given size, and the enumeration and sampling of solutions of a given size.
For each solution space property, we compute it with the same tensor network by replacing the tensor elements with those having an algebra designed for it. % This technique can further be generalized to computing solution space properties beyond the examples given in this paper. % such as the largest several independent sets sizes, the number of independent sets of a given size, and the enumeration of independent sets of a given size.
%With different algebras, the contraction of the same tensor network gives us the different properties.
The different data types introduced in the main text to compute these properties are  summarized in the diagram in \Fig{fig:venn-diagram}. %We go beyond using real numbers and use polyn
The class of problems solvable by a tensor network include but is not limited to maximal independent sets and a variety of other combinatorial problems such as the matching problem, the k-coloring problem, the max-cut problem, and the set packing problem, as detailed in \App{sec:otherproblems}.

%Other than the case with real numbers, we have polynomials and truncated polynomials.
%We combine them with the bit string algebra, the set algebra, and sum-product expression tree algebra for finding, enumerating, and sampling of independent sets at a given size. Three examples are shown to demonstrate how our method can be used to understand the solution space better.
%In the first example, we show how our method can help computing the entropy constant for some hardcore lattice gases on 2D square lattices. We compute the 2D generalization of Fibonacci integer sequence up to size $39$, which is larger than the previous record $37$.
%In the second example, we analyze the pairwise Hamming distances from configurations directly sampled from large independent sets of King's graphs at $0.8$ filling and three regular graphs. We show the strongest numeric evidence so far of the absence of overlap gap property in King's graph and the existence of Overlap Gap properties in three regular graphs.
%In the last example, we show how our algorithm helps people to understand the performance of a quantum variational algorithms on Rydberg atom arrays to find maximum independent sets. This is a research frontier that understanding which can help people design better quantum algorithms to beat its classical counterpart.

It would be interesting to generalize the idea of generic programming to other algorithms for property computing with certain algebraic structures such as those using the inclusion-exclusion principle or subset Onconvolution~\cite{Fomin2013}.
To this end, it is worth mentioning the dynamic programming~\cite{Courcelle1990, Fomin2013} that closely related to a tensor network.
For example, the Viterbi algorithm for finding most probable configuration in the hidden Markov model has a nice matrix product state interpretation~\cite{Han2018}, and the tropical tensor network is very likely equivalent to dynamic programming in solving hard combinatorial optimization problems in certain way~\cite{Liu2021}.
However, while being less restrictive regarding the problem it can solve, e.g. the traveling sales man can be solved easily with dynamic programming but not with a tensor network to the best of our knowledge, it is still an open question whether dynamic programming can have an algebraic interpretation for computing solution space properties beyond finding a single best solution or not.
%Moreover, since the independence polynomial is closely related to the matching polynomial~\cite{Levit2005}, the clique polynomial~\cite{Hoede1994}, and the vertex cover polynomial~\cite{Akbari2013},
%our algorithm to compute the independence polynomial can also be used to compute these graph polynomials.
%By adapting to the style of generic programming, we can implement all the algorithms without much effort.

The source code in Julia language for this paper can be found in our Github repository~\cite{GraphTensorNetworks}. 
There is also a short introduction to this repository as well a gist to show how it works in the Appendix~\ref{sec:technical}.
We expect our tool can be used to understand and study many interesting applications of independent sets and beyond.
We also hope the toolkit we built, including tensor network contraction order optimization and efficient tropical matrix multiplication, can be helpful to other scientific software.

%We also hope this approach of generic tensor networks can inspire future works on tensor network computations.
